Generalized Plackett-Luce Likelihoods

### 2022-09-29 inst/j_sports_econ_F1points.tex
house style for Journal of Sports Economics, issue #181

%% This file is to be used as a template for your submission.
%% Rename this file and replace the text with the text of
%% your manuscript.
%%
%% The standard LaTeX document class "article" is recommended.
%% Use options letterpaper and 12pt.
\documentclass[letterpaper,12pt]{article}

@@

%% Load De Gruyter specific settings
\usepackage{dgjournal}
\usepackage{amsmath}

@@

@@

%% Do NOT include any fronmatter information; including the title, author names,
%% institutes, acknowledgments and title footnotes (author information, funding
%% sources, etc.). Start the document with the first section or paragraph of
%% the article.

@@

\section*{Abstract}
The points scoring system of Formula 1 motor racing is a long-standing
contention among fans and competitors, with changes being
controversial.  The inaugural points system was $(8,6,4,3,2)$, that
is, 8 points to the winner, 6 to second place, and so on and in 2020
the points system was changed to $(25,18,15,12,10,8,6,4,2,1)$.  The
overall ranking of the competitors is determined by summing points
awarded over a season, but this system suffers from two defects:
firstly, the points are intrinsically arbitrary in the sense that the
points system does not affect the competitors' behaviour but can
change the overall ranking.  Secondly, it is difficult to assess
statistical nulls using accumulated points.  Here, I use Plackett-Luce
likelihood to identify a ranking for the competitors, this method
including no arbitrary points system and being amenable to statistical
testing of nulls.  I go on to assess a number of reasonable points
systems objectively, by comparing likelihood-based rankings with
points-based rankings.  In one well-defined statistical sense a Borda
points system is preferred, and in another a ``winner takes all''
system is best.

\section{Introduction}\label{introduction}
Formula 1 motor racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  Season ranking is based on a points
allocation system wherein competitors are awarded points based on race
finishing order; points accumulate additively.  The overall
competition winner is the competitor who accumulates the most points
after the final race.

@@

and this would be a forward ranking Plackett-Luce model in the
terminology of \cite{johnson2020}. A slight generalization allows the
incorporation of non-finishers (DNF etc). If, say, competitors 4 and 5
did not finish, we would have

@@

\section{Formula 1 dataset}\label{formula-1-dataset}

\begin{table}
\centering
%\begin{tabular}{ |c|c|c|c|c|c|  ccc |c| c|c|}
\begin{tabular}{ lccccc  ccc c cc}
 \hline
          &AUS&CHN&BHR&RUS&ESP&MON&$\ldots$&BRA&ABU\\ \hline
Hamilton  &  2&  1&  2&  4&  1&  7&$\ldots$&  4&  2\\
Vettel    &  1&  2&  1&  2&  2&  1&$\ldots$&  1&  3\\
Bottas    &  3&  6&  3&  1&Ret&  4&$\ldots$&  2&  1\\
\raik &  4&  5&  4&  3&Ret&  2&$\ldots$&  3&  4\\
Ricciardo &Ret&  4&  5&Ret&  3&  3&$\ldots$&  6&Ret\\
$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$\\
Hartley   &  0&  0&  0&  0&  0&  0&$\ldots$&Ret& 15\\
Button    &  0&  0&  0&  0&  0&Ret&$\ldots$&  0&  0\\
Resta     &  0&  0&  0&  0&  0&  0&$\ldots$&  0&  0\\ \hline
\end{tabular}
\caption{2017 Season results\label{results2017} table.  Each row is a
  driver and each column (after the first) a venue. We see that
  Hamilton, the first row, came second in Australia, first in China,
  second in Bahrain, fourth in Russia, and so on (Hartley, Button, and
  Resta placed last).  In the first column, we see the result from
  Australia (AUS) in which Hamilton came second, Vettel first, Bottas
  third, and so on.  Here, ``Ret'' means retired and a zero entry
  means ``did not finish''}
\end{table}

@@

\begin{figure}
{\centering \includegraphics{printapiechart-1}}
\caption[Maximum likelihood estimates \label{piechartstrength} of the
  strengths of the top-ranked 11 drivers in Formula 1 motor racing,
  seasons 2016-2019]{Maximum likelihood
  estimates \label{piechartstrength} of the strengths of the
  top-ranked 11 drivers in Formula 1 motor racing, seasons
  2016-2019}\label{fig:printapiechart}
\end{figure}

@@

\begin{figure}
{\centering \includegraphics{likevspoints-1}}
\caption{Ordering of the \label{orderingbypointsandlikelihood}
  top-ranked 11 drivers in Formula 1 seasons 2016-2019.  Horizontal
  axis gives official (points-based) order, and the vertical axis
  gives the likelihood order.  Thus, taking 2017 as an example, the
  points-based ordering would be Hamilton first, then Vettel, then
  \raik; while the likelihood ordering is (reading
  vertically) Hamilton, Vettel, Bottas }\label{fig:likevspoints}
\end{figure}

@@

\begin{figure}
{\centering \includegraphics{likevspointszipf-1}}
\caption[As for
  Figure~\ref{orderingbypointsandlikelihood} \label{orderingbypointsandlikelihood_zipf}
  but points ranking calculated by Zipf's law]{As for
  Figure~\ref{orderingbypointsandlikelihood} \label{orderingbypointsandlikelihood_zipf}
  but points ranking calculated by Zipf's law. Note the generally
  poorer agreement between the two ranking
  systems}\label{fig:likevspointszipf}
\end{figure}

@@

\begin{table}
\centering
\begin{tabular}{ |c|c|c|c|c|}
 \hline
         &    2016 & 2017 & 2018 & 2019\\ \hline
current  &    11   & 1    & 2    & 2   \\
inaugural&     8   & 1    & 2    & 2   \\
Zipf     &     0   & 1    & 2    & 2   \\
Borda    &    11   & 4    & 6    & 2   \\
WTA      &     0   & 1    & 2    & 2   \\
exp2     &     0   & 1    & 2    & 2   \\
 \hline
\end{tabular}
\caption{Degree of agreement for seasons 2016-9 with different
  \label{doapoints} points systems}
\end{table}

@@

\begin{figure}
{\centering \includegraphics{alottaplotsmeancorrect-1}}
\caption[Simulated Formula 1 races, seasons 2016-2019]{Simulated
  Formula 1 races, seasons 2016-2019: mean number of agreeing places
  for each of six points \label{simulated_mean} systems when compared
  with a Plackett-Luce strength
  ordering}\label{fig:alottaplotsmeancorrect}
\end{figure}

\begin{figure}
{\centering \includegraphics{alottaplotswinnercorrect-1} }
\caption[Simulated Formula 1 races, seasons 2016-2019]{Simulated
  Formula 1 races, seasons 2016-2019: probability of identifying the
  Plackett-Luce winner for each of six
  points \label{simulated_winner_correct}
  systems}\label{fig:alottaplotswinnercorrect}
\end{figure}

\begin{figure}
{\centering \includegraphics{alottaplotstotalrankorder-1}}
\caption[Simulated Formula 1 races, seasons 2016-2019]{Simulated
  Formula 1 races, seasons 2016-2019: probability of complete
  agreement between Plackett-Luce ranks and ranks calculated by each
  of six points \label{simulated_complete_agreement}
  systems}\label{fig:alottaplotstotalrankorder}
\end{figure}

@@

\section{Conclusions}

@@

%% BibTeX support
\bibliographystyle{DeGruyter}
\bibliography{formula1}

>>>>

%% This file is to be used as a template for your submission.
%% Rename this file and replace the text with the text of
%% your manuscript.
%%
%% The standard LaTeX document class "article" is recommended.
%% Use options letterpaper and 12pt.
\documentclass[a4,12pt]{article}

@@

%% Load De Gruyter specific settings

\usepackage{amsmath}

@@

\usepackage{setspace}\doublespacing

@@

\begin{titlepage}
  \begin{center}{\Large
      An objective assessment of point scoring systems in Formula 1 motor racing}\\
    Robin K. S. Hankin\\
{\tt hankin.robin@gmail.com}\\
Auckland University of Technology
\end{center}

\end{titlepage}

@@

\newpage
\section*{An objective assessment of point scoring systems in Formula 1 motor racing}
\section*{Abstract}
The points scoring system of Formula 1 motor racing is a long-standing
contention among fans and competitors, with changes being
controversial.  The inaugural points system was $(8,6,4,3,2)$, that
is, 8 points to the winner, 6 to second place, and so on and in 2020
the points system was changed to $(25,18,15,12,10,8,6,4,2,1)$.  The
overall ranking of the competitors is determined by summing points
awarded over a season, but this system suffers from two defects:
firstly, the points are intrinsically arbitrary in the sense that the
points system does not affect the competitors' behaviour but can
change the overall ranking.  Secondly, it is difficult to assess
statistical nulls using accumulated points.  Here, I use Plackett-Luce
likelihood to identify a ranking for the competitors, this method
including no arbitrary points system and being amenable to statistical
testing of nulls.  I go on to assess a number of reasonable points
systems objectively, by comparing likelihood-based rankings with
points-based rankings.  In one well-defined statistical sense a Borda
points system is preferred, and in another a ``winner takes all''
system is best.
\newpage

\section{\centering Introduction}\label{introduction}
Formula 1 motor racing is an important and prestigious motor sport
\citep{codling2017,jenkins2010}.  Season ranking is based on a points
allocation system wherein competitors are awarded points based on race
finishing order; points accumulate additively.  The overall
competition winner is the competitor who accumulates the most points
after the final race.

@@

and this would be a forward ranking Plackett-Luce model in the
terminology of \cite{mollica2014}. A slight generalization allows the
incorporation of non-finishers (DNF etc). If, say, competitors 4 and 5
did not finish, we would have

@@

\section{\centering Formula 1 dataset}\label{formula-1-dataset}

\begin{table}
\centering
%\begin{tabular}{ |c|c|c|c|c|c|  ccc |c| c|c|}
\begin{tabular}{ lccccc  ccc c cc}
 \hline
          &AUS&CHN&BHR&RUS&ESP&MON&$\ldots$&BRA&ABU\\ \hline
Hamilton  &  2&  1&  2&  4&  1&  7&$\ldots$&  4&  2\\
Vettel    &  1&  2&  1&  2&  2&  1&$\ldots$&  1&  3\\
Bottas    &  3&  6&  3&  1&Ret&  4&$\ldots$&  2&  1\\
\raik &  4&  5&  4&  3&Ret&  2&$\ldots$&  3&  4\\
Ricciardo &Ret&  4&  5&Ret&  3&  3&$\ldots$&  6&Ret\\
$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$&$\ldots$\\
Hartley   &  0&  0&  0&  0&  0&  0&$\ldots$&Ret& 15\\
Button    &  0&  0&  0&  0&  0&Ret&$\ldots$&  0&  0\\
Resta     &  0&  0&  0&  0&  0&  0&$\ldots$&  0&  0\\ \hline
\end{tabular}
\caption{\doublespacing 2017 Season results\label{results2017} table.
  Each row is a driver and each column (after the first) a venue. We
  see that Hamilton, the first row, came second in Australia, first in
  China, second in Bahrain, fourth in Russia, and so on (Hartley,
  Button, and Resta placed last).  In the first column, we see the
  result from Australia (AUS) in which Hamilton came second, Vettel
  first, Bottas third, and so on.  Here, ``Ret'' means retired and a
  zero entry means ``did not finish''}
\end{table}

@@

\begin{figure}
{\centering \includegraphics{printapiechart-1}}
\caption[\doublespacing Maximum likelihood estimates \label{piechartstrength} of the
  strengths of the top-ranked 11 drivers in Formula 1 motor racing,
  seasons 2016-2019]{\doublespacing Maximum likelihood
  estimates \label{piechartstrength} of the strengths of the
  top-ranked 11 drivers in Formula 1 motor racing, seasons
  2016-2019}\label{fig:printapiechart}
\end{figure}

@@

\begin{figure}
{\centering \includegraphics{likevspoints-1}}
\caption{\doublespacing Ordering of
  the \label{orderingbypointsandlikelihood} top-ranked 11 drivers in
  Formula 1 seasons 2016-2019.  Horizontal axis gives official
  (points-based) order, and the vertical axis gives the likelihood
  order.  Thus, taking 2017 as an example, the points-based ordering
  would be Hamilton first, then Vettel, then \raik; while the
  likelihood ordering is (reading vertically) Hamilton, Vettel, Bottas
}\label{fig:likevspoints}
\end{figure}

@@

\begin{figure}
{\centering \includegraphics{likevspointszipf-1}}
\caption[\doublespacing As for
  Figure~\ref{orderingbypointsandlikelihood} \label{orderingbypointsandlikelihood_zipf}
  but points ranking calculated by Zipf's law]{\doublespacing As for
  Figure~\ref{orderingbypointsandlikelihood} \label{orderingbypointsandlikelihood_zipf}
  but points ranking calculated by Zipf's law. Note the generally
  poorer agreement between the two ranking
  systems}\label{fig:likevspointszipf}
\end{figure}

@@

\begin{table}
\centering
\begin{tabular}{ |c|c|c|c|c|}
 \hline
         &    2016 & 2017 & 2018 & 2019\\ \hline
current  &    11   & 1    & 2    & 2   \\
inaugural&     8   & 1    & 2    & 2   \\
Zipf     &     0   & 1    & 2    & 2   \\
Borda    &    11   & 4    & 6    & 2   \\
WTA      &     0   & 1    & 2    & 2   \\
exp2     &     0   & 1    & 2    & 2   \\
 \hline
\end{tabular}
\caption{\doublespacing Degree of agreement for seasons 2016-9 with
  different \label{doapoints} points systems}
\end{table}

@@

\begin{figure}
{\centering \includegraphics{alottaplotsmeancorrect-1}}
\caption[\doublespacing Simulated Formula 1 races, seasons
  2016-2019]{\doublespacing Simulated Formula 1 races, seasons
  2016-2019: mean number of agreeing places for each of six
  points \label{simulated_mean} systems when compared with a
  Plackett-Luce strength ordering}\label{fig:alottaplotsmeancorrect}
\end{figure}

\begin{figure}
{\centering \includegraphics{alottaplotswinnercorrect-1} }
\caption[\doublespacing Simulated Formula 1 races, seasons
  2016-2019]{\doublespacing Simulated Formula 1 races, seasons
  2016-2019: probability of identifying the Plackett-Luce winner for
  each of six points \label{simulated_winner_correct}
  systems}\label{fig:alottaplotswinnercorrect}
\end{figure}

\begin{figure}
{\centering \includegraphics{alottaplotstotalrankorder-1}}
\caption[\doublespacing Simulated Formula 1 races, seasons
  2016-2019]{\doublespacing Simulated Formula 1 races, seasons
  2016-2019: probability of complete agreement between Plackett-Luce
  ranks and ranks calculated by each of six
  points \label{simulated_complete_agreement}
  systems}\label{fig:alottaplotstotalrankorder}
\end{figure}

@@

\section{\centering Conclusions}

@@

%% BibTeX support
\bibliographystyle{apalike2}
\bibliography{formula1}
